---
title: "Experiments in generating audiobook files with epub2tts-edge"
author: "John C Nash"
date: "2025-01-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Initial version 2024-07-29, revised 2024-10-19, 2025-01-09

## Background

Audiobooks have become a popular way to hear rather than read the content of
books, in addition to being an important resource for the visually impaired.
Unfortunately, creating good-quality audiobook files is not easy. Written content
has many features of typography, layout, illustration and idiosynchratic expression
that do not carry over to the spoken word. Moreover, most 
readers, even professional actors,  have difficulty maintaining their enunciation and 
providing expressiveness over the generally extended duration of a book.
Furthermore, the services of professional actors and readers are expensive relative
to the resources of the visually impaired and amateur authors.

The development of large language models for converting text to speech (i.e. TTS)
has reached the stage where it could be considered a possibility for carrying 
out the task of generating audiobooks from textual form, for example, epub 
files. This article is a set of notes on some experiments with freely available
software to do precisely this.

Popularly such software is called "artificial intelligence", but "large language
models" more correctly describes the programs, since they use models of how 
language is pronounced and expressed to create the sounds of a human voice in a
particular accent and style. In particular, the software used
in developing this report comes from the Github project **epub2tts-edge** of
Christopher Aedo. See https://github.com/aedocw/epub2tts-edge. Aedo has done a
great service in this and the earlier **epub2tts** project which is based on 
the Coqui AI TTS tools (https://github.com/coqui-ai/TTS). Moreover, he has 
responded quickly and helpfully to queries.

There are undoubtedly other tools that I could have used, but experimentation is
very time consuming, and preliminary experiments showed promise, first with 
**epub2tts**, then somewhat better results with **epub2tts-edge**. This led to
more, and more ambitious, experiments, as well as some investigations into ways
to improve outputs or make them more usable to potential listeners.

## Source works

I have generally put my creative writing on the UK writers' site *https://obooko.com.*
This currently is about a dozen novels and two collections of memoir/short story
material. A feature of several of my works is that they involve more than one 
language. My first novel, **Thursday Afternoon**, for example, has a scene where
a young Canadian airforceman is working with some British soldiers and a Flemish
woman to find and clear a landmine in 1944 Flanders. The Flemish woman communicates
with the Canadian in French, but falls back into Flemish on occasions. There are
thus three languages mixed in a couple of pages of text, which is a challenge to
the software. Moreover, I have yet to discover how to inform the software of the
language to be read, though I feel certain there must be a way to do this. For 
example, I am surprised that simple tags cannot be put into the text to be read
giving the language which follows. Similarly, some indication where emphasis is
desired would be useful.

## Installation of the software

Setup of **epub2tts-edge** followed instructions in the github repositories to 
install the software for the two 
projects. Note that this may involve a considerable download of code and
binaries depending on the existing software installation. I worked with Linux Mint 
version 22 (Wilma). One limitation for me is a lack of
CUDA-capable graphical processing units (GPUs), so performance of the
tools on my computers was somewhat lacklustre. Typically rendering of
an epub to an m4b or mp3 took over an hour.

Note that the `epub2tts-edge` software currently calls for Python support using
at least version 3.11 or later. The audio rendering also needs an active internet
connection. 

The software was apparently developed for a Linux environment. While there is the possibility
in Windows of running the Windows Subsystem for Linux, I have chosen to use Oracle
VirtualBox to install Linux Mint 22 (Wilma) and run the software within a virtual
machine (VM). This has the advantage of isolating the text to speech work from other ongoing 
activities. Initially, my working platform was an earlier Linux Mint with Python 3.09.
I did not wish to disturb some stable facilities of my working environment, 
though I have since upgraded the host
to LM22. My writing colleague and coordinator of the Stittsville Creative Writers Group, 
R J Partington, installed LM22 under Windows. Another colleague in the group, 
Allan McCarville wanted to install on a Macintosh, but we had difficulty getting
that set up, and it seems likely he will repurpose an older PC to work with audiobooks.

## Standard usage of the software

We assume we have followed instructions in the Github repository to install the
software. 

**epub2tts-edge** is intended to be applied first to an **epub** file. Such files
are described in https://www.w3.org/publishing/epub3/ and 
https://en.wikipedia.org/wiki/EPUB; they are packaged collections of HTML text 
and metadata. Suppose that
we have a file **mybook.epub**. Then we first convert this file to a properly
structured plain text file **mybook.txt** by opening a terminal (command line
screen) and issuing the command

```
       epub2tts-edge mybook.epub
```

This takes only a few seconds to execute. 
This (usually) creates a file mybook.txt and extracts a cover image 
mybook.png from the epub file. The text file begins with two 
lines of the form

```
Title: Dodging the Potholes
Author: John C. Nash
```

It also has some lines that commence with a hash symbol (#) that 
give "chapters" or "sections" and the text behind the hash is read
out in the output audio as a chapter heading. However, the algorithm
by which these lines are created is not clear to me, but they can be
manually inserted into the file or edited with a plain text editor.
For example, generated ones like

```
# Part 24 
```

can be edited to a more desirable phrase like 

```
# Part 24 -- At the party
```


Text following the hash is read
as the title of the section. Below we discuss how to break up the work and have separate
audio files, which is important in saving time and effort for correction of individual parts
of the overall novel or story.

### Failure of the epub to text process

We have observed some cases where `epub2tts-edge` fails to carry out the conversion
to text and extraction of the cover image. There are at least two other approaches 
we have used, and some others we have investigated:

1) **calibre** program `ebook-convert`

This tool is part of the **calibre** ebook management software. We simply use the
command

```
ebook-convert myfile.epub myfile.txt
```

Generally `myfile.txt` will be more or less correctly formatted, though we have
noticed that the initial "Title:" and "Author:" may be combined in one line. This
can, of course, be corrected with a text editor. This tools should be the first 
alternative to `epub2tts-edge` when that
program fails to convert the epub successfully.

2) **pandoc**

```
pandoc -f epub -t plain -o myfile.txt myfile.epub
```

will create a text version of the epub. Unfortunately, it often breaks paragraphs
into multiple lines which cause undesirable pauses in the eventual audio file(s).
(See 
https://unix.stackexchange.com/questions/647686/convert-epub-to-txt-and-preserve-original-formatting).
Moreover, the section delimiters ("# (some text)") are often missing. 
`pandoc` also leaves two line endings between "paragraphs". It should be 
noted that epub2tts-edge then pauses for each line ending, so these single 
line endings need to be removed. 
However, there are some parts of the epub such as the title and cover pages where we would like 
to retain the single line endings (or double them to ensure separation of important fields that 
should be read separately).
Note that omitting the "-t plain" option leaves some tags within the resulting text file that 
could cause unwanted spoken material. 

3) Some word processing programs like **Word** or **LibreOffice** may be able to 
save a text file, but likely the user will need to manually edit the output.

CAUTION: Plain text often gets corrupted (or perhaps "enhanced") by word processing
software. It is advisable to use an editor for plain text. In Linux, I like 
**L3afpad** and **gedit**. In Windows, the rather ancient **Notepad** is still
a reliable standby. Below we discuss some tweaks to the text file that are generally
needed to give a polished audiobook. For example, each paragraph should be a single
line terminated with a line ending.

4) I could not get the software **epub2txt2** (https://github.com/kevinboone/epub2txt2) to 
compile (and hence was unable to install it).

5) **epub2txt** is a Python program (https://github.com/ffreemt/epub2txt) that can be
easily installed with the command `pip install epub2txt` if Python is properly set
up. This seems to get all the material, but gives no spaces between what would be
"paragraphs". Moreover, verbatim material has line endings, so some attention to 
that detail may be relevant.

### Conversion to audio file(s)

The much longer process that **epub2tts-edge** carries out is converting the 
text file to an audio file of type
m4b (https://www.lifewire.com/m4b-file-2621958). For a novel, this may take of the 
order of an hour if the work is
in a single computer text file. Breaking up the text file into parts is therefore
helpful for repairing glitches in individual sections, and my experience is that
these glitches are inevitable. We want to avoid having to re-run the process for
the whole book rather than one section.

The second stage run of **epub2tts-edge** generally requires us to specify a "voice" 
or "speaker". If no speaker is given, a neutral English-only speaker with a quite
moderate American accent is used, which can be specified by the identifier **en-US-AndrewNeural**.
Moreover, for multilingual texts we need to use one of a few special multilingual voices.
For example, the specifier **en-US-AndrewMultilingualNeural** will use the same voice
as the default **for English**, but process multiple languages using **different** speaker
voices for the other languages, which is disconcerting for the listener. Note that it seems 
that different "speaker" specifications will handle the same text differently and some 
words pronounced correctly by one voice will be mangled by another. We have yet to
find out how to get a seemingly common voice for different languages.

The second stage command can also specify a cover image which can be in jpg or png format.
The command is

```
    epub2tts-edge --speaker speakerID --cover mybook.png  mybook.txt
```

where `speakerID` is a valid speaker identifier and `mybook.png` (or some other png
or jpg image file) 
is a cover image available in the same directory as `mybook.txt`. The output will be
a file `mybook-speakerID.m4b`. This can be a very large file of several hundred megabytes
for a full novel, and take 8-16 hours to play. The m4b format allows bookmarks so the
user can pause and restart. On the other hand, older devices may not be able to play
these files. They are, however, convertible to other audio file formats, in particular
mp3. 

As mentioned, paragraphs should not have line breaks. Such line breaks in the text 
file `mybook.txt` cause quite long pauses in the
voice reading. Therefore line breaks that are non-pausing should be removed, giving
full paragraphs on a single "line". 

## Some important general problems

### Omitted text

**WARNING**: I have discovered that text in an epub file that is in a monospaced
(i.e., typewriter-like) font is somehow eliminated completely from the file `mybook.txt`
and must be re-inserted manually. This is a nuisance and can cause grief. For example,
in my novels, I use such fonts for "letters" and other correspondence or quotations
such as the text on a tombstone. I have had some email correspondence with Christopher
Aedo about this, but so far we have not found the source of the glitch. (Note 2024-11-01:
This may have been fixed, at least in part, but has not been tested.)

As mentioned, it is feasible to replace the first stage run of epub2tts-edge with a different 
translator from epub to text. Some other choices were noted above.

I have not tested to see if possibly material in some other fonts may be omitted.

### Lost features

I have not been able to get strikeout, bold or italic to be expressed in 
the audio output.

Emphasis, by way of capitalization, is also lost, as can be seen by processing the file 
**cap4emphasis.txt** (the text below) through epub2tts-edge. Note line endings have been removed.

```
# Capitals for emphasis

Putting words in capitals to emphasize them will only get them pronounced letter by letter in some cases. Hence capitalized "ARE" is not emphasized. Separated letters gives A R E. Simple "are" does not get emphasized. It would be useful to be able to ask **epub2tts-edge** to give emphasis.

"I said," often has awkward pauses or mis-emphasis.
```

### Difficulties with language change

When the language of the text changes, particularly with dialogue, there can be 
unwelcome results:

- Sometimes the language model, which appears to be linked into the speaker voice chosen,
either fails to recognize the language change, so continues in the current language
setting inappropriately, or else will continue after the "new" language in that language
when it is not correct.

- The voice of the speaker may change drastically in the new language, which is 
disconcerting for the reader.

I have not (at 2024-1-9) yet found work-arounds for these issues.


## Workarounds

Some glitches in pronunciation can be overcome by use of phonetic text. Here
are a number of examples, all of which are small files that can be tested using

```
  epub2tts-edge [file].txt
```

where `[file]` is replaced by the appropriate identifier. Note that in some cases
the long lines needed by epub2tts-edge give display problems for the Rmarkdown
editor in R-studio (https://www.rstudio.com/categories/rstudio-ide/) 
and the resulting pdf output in this document.


### 0pauses.txt

```
0pauses.txt

Various things cause pauses.

Line endings.

Like
these
ones.

A period will give a break. Just like that. But shorter than a line ending.

And John C. Nash sounds incorrect compared to John C Nash with no period
behind the middle initial.

But dashes -- seem to be ignored.

What about a semicolon; does that work? More or less. Perhaps dashes are same.
```

### titles.txt

```
# titles.txt

Abbreviations may or may not be expanded.

Lt. Cmdr. Try full words "Lieutenant Commander" or phonetically Leftenant Commander

Putting M I 9 together gets MI9.  Better separated.

Ms. (not sounding like Miz). Try Miz.

Mrs. could be better MISSUS or Missus. Note that epub2tts-edge seems to get this to work OK.

We can also try versus and vs.

Monsieur Lefebvre not M. Lefebvre

POWs  ("s" not pronounced sometimes). Try P O W's. 
```

### names.txt

Proper names, especially those originating outside English, cause mispronunciation.

```
# names.txt

Vaudreuil try  vohdroy
A / C or Ae slash C for air conditioning or alternating current. Spell it out.
Arnprior Arnpryor Arnprye-ore
Belisle  Belleel
Jean Godin use zjonn gohdann
Lebreton Lebretton
Levis try Levee
Long Sault  (pronunciation)  Long Soo
saltpetre  saltpeter
Paule try Pohl

Almonte use Almont
arboretum
Aylmer
Boucherville probably OK
Dieppe 
Domenico Nacci or Nachi
Erminia
escapees
escheat
Frieda
Greenock -- seems OK
I-81 
Jocelyn
kinesiology and phys. ed. or phonetic fiz ed
Leica 
Maniac
Model A Ford
Myrna pronounced strangely by female voice default of epub2tts
Ontario
Rideau pronounced Rydoh in epub2tts
Rimouski
Tremblay mispronounced  try T R O M B L A Y  Tromblay
V I M Y or Vimy pronounced Veemy or Vy me? By using V I M M Y we get Vimmy.
```


### homonymns.txt

```
# homonyms.txt

"read" sometimes reed, not red

bough of a tree, bow of a ship, a ribbon in a bow.

bowed down

learned vs learnt

row -- pronounced as rou not roe

lead pronounced as "leed" (led for weight)
```

### numberpunctuation.txt

```
# numberpunctuation.txt

Numbers, units and punctuation give difficulties.

2/3 versus two thirds

.2 or point 2 is just 2 with pause. Not Point 2

Brackets (), [], and {} are not handled well

web addresses are often mangled

Lower case M M may be millimeters. M M M is mmm.
35mm vs 35 mm
No. 2  not number 2
cwt. versus hundredweight
```

### other.txt

```
# other.txt

en-US-AndrewNeural speaker handles these correctly
escapees
steno for stenographer
`unmaking` not pronounced correctly in epub2tts
```

### Dialect

Sometimes we want to write dialogue with the speaker using local pronunciation.
Particular cases may be the dropping of the letter H by some working-class
British speakers. This can give problems, and need phonetic replacement.

```
# dialect.txt

I'll 'ave the recording of Ave Maria.

I'll avv the recording of Ave Maria.
```

### Vocalizations

Vocalizations like "oh", "er", "hmmph", "hmmm" sometimes get mangled, though 
epub2tts-edge with the en-US-AndrewNeural voice seems to work reasonably well. 

```
# vocalizations.txt

epub2tts-edge does not seem to know how to properly enunciate some vocalizations common in human speech.

The surprise expressive "oh" is often mis-said. "Oh, I really must fix that."

Disdain, as with "hmmph". Hesitation as with -- er -- and -- hmmm -- or realization with "ah" are potentially problematic.

```

## Audio post-processing

It is sometimes necessary to post-process the audio files.

### Conversion to mp3

Since many older devices cannot play m4b files, but are intended for mp3 files, 
we may need to convert. Unfortunately, most such devices do NOT allow the user
to save the position of a paused play. Therefore it is useful to have an audiobook
in mp3 form in multiple smaller files. One way to do this is to build the audiobook
in pieces, and this is the recommended approach. Nevertheless, **epub2tts-edge**
produces files as m4b files, and these need conversion. I do this with **ffmpeg**
and batch process them with a small (Linux / bash) command-line script. 

```
#!/bin/bash
#
# convm4b
mkdir newfiles
for f in *.m4b; do 
ffmpeg -i "$f" -codec:v copy -codec:a libmp3lame -q:a 2 newfiles/"${f%.m4b}.mp3"
done
```

**Open question** What are Windows / Mac equivalents?

### Breaking a large mp3 file into pieces

I have also had to break
up a large file (converted from the m4b of an entire novel). For this I used the
**audacity** software which is available on most common platforms. This is a graphical
user interface program allowing for sections of an audio file to be cut and pasted
and exported in various formats, including mp3. 

### Concatenating audio files

Sometimes the mp3 files are really too small, so the user needs to keep loading
them. In that case we want to concatenate several files. **ffmpeg** can do this
using a text file containing a list of the files to be processed in the order they appear.
An example command is 

```
ffmpeg -f concat -i mylist.txt -c copy output.wav
```

which will produce a file of type wav. 

File mylist.txt must have lines of the form 

```
file 'myfile.type' 
```

where type is one of acceptable types to ffmpeg, which does not include
m4b unfortunately, even though we used ffmpeg above to do conversions.

Another tool is **mp3wrap**, which uses a command

```
mp3wrap [options]  OUTPUTFILE  f1.mp3  f2.mp3  [f3.mp3]
```

Unfortunately, there seems to be a limit to the number of files that can be included,
but the number does not appear to be documented. One also needs to be careful to 
ensure the ordering of the files.

I was unable to get the program **mp3cat** from  https://github.com/dmulholl/mp3cat
to work at all. 

### m4b-tool

https://github.com/sandreas/m4b-tool claims to do many of the operations we might
wish to do to render audiobooks in convenient sizes etc. Unfortunately, I found

- the software was NOT easy to install. 

- I needed to get the `docker` software, 
then try to install m4b-tool instance, 

- I needed to create an alias to run the software (which disappeared
over reboots, so I needed to explicitly save to `.bashrc`). 

- I needed to add myself to the docker group before the tool would run.

- Once installed, it did NOT work in all directories e.g., the shared
directory of the Linux Mint 22 virtual machine in which I run all audiobook 
work to keep software intact.

- Once I did get it to start running, I got sporadic errors depending on where
I started the software relative to input and output locations. 

- It seems the software is fragile concerning cover images. 

- I did manage to get the software to "sort of" work, with reported errors,
though the output was satisfactory.


### Changing speed or tempo of audio

**audacity** allows the speed or tempo (pitch may change with one of these) to
be altered. For some texts and some "speakers" (i.e., voices) a slowdown of 
approximately 8% made the material easier to comprehend.


## Special Tools

To process the main source text for the second-stage of epub2tts-edge, I have 
written several small scripts to perform particular tasks. I have written these
in **R** or in the Linux **bash** script, but those choices were purely for
convenience. The tasks are relatively straightforward, but tedious to carry out
without the automation.

### Section numbering

Most of my books do not have regular chapters, but do have sections. 
These are typically, but not always, delimited by "* * *".
I use a text editor and replace these (and any other suitable breakpoints)
with  

```
# Part ??
```

An R script numpart.R (see Appendices) is then used (after editing filename specifications
within the script) to auto-replace the "??" with suitable numbers. Note that we
want these numbers to all be the same length for processing to audio files, 
so leading zeros are strongly recommended in the numbering, and the script
prepares such numbers.

The section dividers provided by epub2tts-edge can also be edited
to provide useful output. I have used lines like

```
# The title of this book
```

or 

```
# Capsule comment
```

with the latter for the "teaser" comment for the book.

### Section splitting

It is useful to have individual files of text for each section of a book to
convert to audiobook form. Such files are smaller and more quickly processed, 
so errors or glitches can be repaired and the files re-processed. File 
**splitnovel.R** in the Appendices does this.

Notes:

- we may wish to consolidate several sections after splitting so the audio
playback files are of a reasonable size. This can be done with script **combpart.R**
in the Appendices.  Titles for the consolidated sections could be titled withing 
the text source for conversion by adding title lines in the form

```
# Module A of My Excellent Book
```


- there are issues with sections that are named versus 'Part 043' style, in that
mixing these can be awkward. A suggestion is to use a style

```
# Part 043 -- The plot thickens
```


## A recipe

### 1) Extracting the cover and the text

`epub2tts-edge mybook.epub`

should give the cover image in file `mybook.png` and "text" 
in file `mybook.txt`. 

The program **epub-cover** can be used to extract the cover image. This is 
part of the Ruby package epub-parser from https://gitlab.com/KitaitiMakoto/epub-parser.
The syntax 

`epub-cover mybook.epub` 

will output whatever cover file is embedded in the epub, which may have a name
somewhat different from the epub title, and may be a JPEG file. Note that one
may choose to use epub2tts-edge to get mybook.png and then ignore the text file
produced in case it is missing material that was in monospaced font.

Note that the `ebook-edit` of the **Calibre** ebook software or `sigil` can both
export the cover image.

### 2) Add Title and Author

The first lines of `mybook.txt`, which can be copied to a file 
`mybookheader.txt`, should be something like

```
Title: My fabulous book
Author: Me A. Writer
```

Moreover, the software will try to create "parts" denoted with the hash (#) symbol
followed by a title, but this will depend to a large extent on the epub content
structure.

### 3) Check for monospaced material and insert

Unfortunately, as we have noted, **epub2tts-edge**, using the Python library 
**Beautiful Soup** (according to Christopher Aedo), fails to extract
any material in a monospaced font.

**ebook-convert** does the job reasonably well, but we may need to ensure the
Title/Author header of the type `mybookheader.txt` mentioned above is present. 
Moreover, some pauses in the address and date
on letters may need to be inserted as punctuation or line endings.

The issue of material which may be dropped by epub to text conversion is, I 
believe, likely to remain for some time. This means that users/authors still need
to check the output of the processes in the epub to audio conversion.

### 4) Consolidate paragraphs

- Run the program **novelparas.R** to ensure text paragraphs are consolidated
for the audio output. Even if the text has been extracted with epub2tts-edge,
there may be manually inserted monospace text that requires paragraph consolidation
to remove unwanted line endings.

### 5) Edit for section divisions and their titles

Whatever divisions are present (I divide sections with `* * *`) can be edited to 
the form `# Part ??`, or such divisions can be inserted. Blank lines above
and below the division are recommended. This can easily be done with most text editors.

### 6) Section numbering

Run **numpart.R** to get sections numbered. This replaces the `??` in `# Part ??`
with appropriately sequenced numbers.

### 7) Standard cleanup of the text file

Besides the Author/Title header and the monospaced material, it is useful to check
the text (before splitting by sections) to fix some issues.

- **Front matter** often has symbols, such as copyright, or Internet addresses for
web pages or email. These may need attention to get them read in a comprehensible
way by the artificial voice. 

- Remove periods after initials in author name(s) to avoid unnecessary pause.

- Remove the copyright symbol (©), if necessary expanding to the word "copyright".

- Put spaces in email addresses e.g. `j.smith@isp.org` becomes `j . s m i t h @ i s p . o r g`.
  Similarly expand abbreviations and postal codes so they are not read as if they are words.

- Expand province or state abbreviations. 

- Some paragraphs as text will not offend the book reader, but when in audio form
will not have the right pauses, so need splitting or combining
for better flow when read out loud.

- Check that there is a line ending for the final line of the mybook.txt file.
This seemingly trivial change bit me in one conversion effort.

- Add section dividers for the title page(s), the capsule comment and any
preamble. Optionally expand or diminish the section divisions.


### 8) Phonetic corrections

Proper names, dialect, homonyms and similar issues for text to speech should
be corrected. To do this, I wrote a small program **wordsub.R** as a prototype
for this operation. This program
needs a set of source / replacement choices. We shall suppose this is in the
form of lines

```
old (tab)  new
```

This assumes 1-1 choices. We will manually deal with situations where there is 
more than one choice using a text editor. **wordsub.R** uses a named file 
**phoneticsub.txt** which has entries as described
above. To save computing time, it is worthwhile having a "master" file, call
is **phoneticsub.full.txt** with all possible replacements. This file can be
copied to **phoneticsub.txt** and unnecessary entry lines simply removed.


### 9) Section splitting

Run **splitnovel.R** to divide the source text into individual sections. This is
helpful later when or if we discover that the conversion of text to speech has
made an error or glitch. We can then attempt to correct the issue for just the
single (small?) section. Once the text is adjusted appropriately, we can then re-work
the section, if necessary inserting it into a consolidated module where it belongs.

### 10) (Optional) Text section consolidation

To provide suitably sized audio files, we could consolidate the text sections into modules
using **combpart.R**. 

However, it is likely more human-efficient to create small audio files then 
put these together with the help of a suitable program such as `joinmp3.R`.

### 11) Conversion to speech

The script **runedge.R** is intended to process the text modules (or sections)
and produce files of type m4b. Note that this script is currently set up to be
first edited to provide the right filename root for the modules and the
correct cover image. 

This stage of the process takes quite a long time -- typically of the order of
an hour. 

### 12) Conversion to mp3

Unfortunately, the tools for consolidation of sets of small m4b files into
larger blocks are not, as far as we have so far found at least, reliable.
There is a tool `m4b-tool` from https://github.com/sandreas/m4b-tool, but
we could not get a satisfactory result. 

The `ffmpeg` program does offer concatenation of mp3 files, and also will
convert m4b files to mp3 (though apparently not concatenate m4b files).
We set this operation up as a script `convm4b.sh` which is given in an
appendix. (Note that the scripts may be still under development.)

### 13) Consolidation of audio blocks

We want our finalized audiobooks to be in parts small enough for users
to navigate, that is, to move to particular parts. Our experience is that
this means blocks of around one to two hours, or between five and ten 
blocks for a book of reasonable length. With files of type m4b users can
generally, with appropriate player software or devices, move quickly between
sections. Still, we believe division of audiobooks into modestly-sized
blocks of audio time is sensible.

The program `joinmp3.R` calculates which small mp3 files to combine, then
combines them, and then does some other operations as described in subsequent
steps.

### 14) Convert consolidated mp3 blocks to m4b format

`joinmp3.R` also will convert the consolidated mp3 audio files to m4b
format. It uses `ffmpeg` to do this. 

### 15) Add a cover page to the files

`joinmp3.R` will also add cover images to each of the large block files,
both mp3 and m4b.

The program **AtomicParsley** (https://github.com/wez/atomicparsley) can be
used to insert a different cover image into m4b (and some other audio) files.
One syntax is

```
cp myfile.m4b tmp.m4b
AtomicParsley tmp.m4b --artwork REMOVE_ALL
AtomicParsley tmp.m4b --artwork cover.jpg -o tmpx.m4b
mv tmpx.m4b myfile.m4b
rm tmp.m4b
```

However, we carefully avoid having an image in the mp3 blocks, so can
simply add the "--artwork" and also author metadata (with the "--artist" 
tag). 

Atomic Parsley does not, unfortunately, add images to mp3 files, and 
we do this (in `joinmp3.R`) with `ffmpeg`. It is important to note that
this NOT be performed before conversion of the mp3 files to m4b. We spent
the better part of an afternoon trying different methods to insert the 
cover image and achieve the appropriate mp3 to m4b conversion before a
careful test established this.

There are some other tools useful for removing or inserting images, of 
which the `eyed3` package is useful (with command `eyeD3`).

### 16) (Optionally) Adjust audio file characteristics

The program **audacity** is useful to change speed, tempo or pitch of audio
files. Some text-to-speech models create output that is very quick. Slower
and lower-pitched output may be helpful to listeners. We found an approximate
8 percent slowing of playback (with concurrent pitch reduction) was helpful
with some of the voices.

## Open needs

- The loss of material in the initial conversion from epub to text is a 
serious issue. A fix for this first stage of epub2tts-edge so that monospaced 
(or other problematic) material is always included. An alternative would be 
tools to work with other epub to text translation tools such as **pandoc** or 
**epub2txt** mentioned above that also establish the paragraph structure
without line endings. At the moment a more or less plain text editor is used
for this work.

- Until and unless the large language models can cope with various proper 
names, homographs and dialect, tools to streamline the replacement of selected
words with phonetic substitutes will be welcome. It would be helpful to have a
GUI editor that highlights target words and offers replacements. Such tools
already exist for spelling checkers, so this is not an idle possibility.

- epub2tts-edge needs an active internet connection to function. There are a
number of situations where it would be important to be able to prepare audio
output from text WITHOUT the need for a connection to the Microsoft service.
Moreover, it is by no means certain that this service will be continued, or
continued on the same basis as at present.

## Conclusions

This article underlines the reality that we are still in the early stages of
large-scale text to speech, at least in the open-source world. 

- We see a large number of small but important "glitches" in the text-to-speech
engines. While not mentioned much in the article, the Coqui AI TTS was tried,
but found wanting.

- Workarounds have been found for most of the glitches discovered. These have,
however, requires quite a lot of work. 

- Pronunciation and emphasis issues still are found. These are not major, but may
still sound jarring to the listener.

- The lack of a way to tell the TTS engine what language a segment of text should 
be used is a serious fault and causes a number of issues. It would also be useful
to be able to specify emphasized or softer speech, or to specify the "speaker"
voice for novels with more than one narrator, such as 
https://www.obooko.com/free-historical-fiction-books/dodging-the-potholes .

- Some books require multiple voices. This is relatively easy to arrange if the
text blocks are broken up, but tags to allow the "speaker" to be changed would
be helpful.

- The **edge** TTS engine require the computer processing text to be internet-
connected. This is a serious obstacle for some applications. It also suggests 
that users may be vulnerable to being held hostage by Microsoft should they
suddenly impose a charging model on the parts of edge-TTS now open.


## Appendix 1: epub2tts-edge voices

There is a list of voices at the end of this document. This was obtained
using a feature of the `edge-tts` project from https://github.com/rany2/edge-tts. This can be installed under Python 3.12
in the Linux Mint 22 Wilma operating system with the terminal command

```
pipx install edge-tts
```

Then the command

```
edge-tts --list-voices > edgevoices.txt
```

will create a text file of documented voices.

### How to get audio samples

The program **get-edge-tts-samples.py** shows how to get samples of audio for
different edge-tts speakers.

```
#!/usr/bin/env python3

"""
Basic example of edge_tts usage.
"""

import asyncio
import edge_tts

NAMES = ['en-US-AnaNeural', 'en-US-AndrewNeural', 'en-US-AriaNeural', 'en-US-AvaNeural', \
 'en-US-BrianNeural', 'en-US-ChristopherNeural', 'en-US-EmmaNeural', 'en-US-EricNeural', \
 'en-US-GuyNeural', 'en-US-JennyNeural', 'en-US-MichelleNeural', 'en-US-RogerNeural', \
 'en-US-SteffanNeural', 'de-DE-FlorianMultilingualNeural', 'de-DE-SeraphinaMultilingualNeural',\
 'en-US-AndrewMultilingualNeural', 'en-US-AvaMultilingualNeural', 'en-US-BrianMultilingualNeural',\
 'en-US-EmmaMultilingualNeural', 'fr-FR-RemyMultilingualNeural', 'fr-FR-VivienneMultilingualNeural']

TEXT = "This is a longer string that I am sending to text to speech, and using the python module directly."

async def amain() -> None:
    """Main function"""
    for name in NAMES:
        output = f"sample-{name}.wav"
        print(f"Saving {output}")
        communicate = edge_tts.Communicate(TEXT, name)
        await communicate.save(output)


if __name__ == "__main__":
    loop = asyncio.get_event_loop_policy().get_event_loop()
    try:
        loop.run_until_complete(amain())
    finally:
        loop.close()

```

There are a number of lists of edge-tts voices. One is at https://github.com/rany2/edge-tts/blob/master/README.md


## Appendix A: Possible troubles

The following text is believed to be likely to give problems so that 
audio output will need to be checked. Different voices appear to have
different language models, so some will "work" while others will not.

```
# Oddbods.txt

Nene -- This is an early Rolls Royce jet engine using a centrifugal complressor of the
         Whittle style. It was used in the De Havilland Vampire and a few other early
         jet aircraft.  Better to use Neen to get the right pronunciation.
Sterling currency expression 17/6 is better written 17 and 6. 
ha'penny  as hah penny, not haypenny
"wan" --> W A N 
Marthe -- more like Martha than Mart
Feltham  "ham"
French mis-pronounced. "687 25th Avenue"
Paule, Lise (more like Liss)
souper  
“Sacrément!”
dommage
compris
cintres
“C’est un très grand réfrigérateur.”
tarte au sucre
erable
Houle
Decarie
Ste Catherine
Outrement
Ultramontes
Universit'e
maudit
Anges
Sinterklaas
Bolduc
Pieds nus dans l’aube
Sault Ste Marie,   Regina
Metis
Mon Dieu
V-1 may be better as V 1
```

## Appendix B: Software tools associated with this article

### convm4b.sh -- convert m4b files to mp3

```{r comment='convm4b.sh', echo=FALSE}
cat(readLines('../baseabsw/convm4b.sh'), sep = '\n')
```

### combpart.R -- combine individual book sections 

The purpose of this script is to create texts that will have a reasonable 
audio file duration, particularly when in mp3 form, which is more difficult in most cases to position or resume playing. The number of parts to combine is 
set at `np <- 10` in the present version.

```{r comment='combpart.R', echo=FALSE}
cat(readLines('../baseabsw/combpart.R'), sep = '\n')
```

### novelparas.R -- remove single line endings to get paragraphs

**epub2tts-edge** will pause on line endings. Paragraphs are therefore assumed
to be terminated by a line ending.

```{r comment='novelparas.R', echo=FALSE}
cat(readLines('../baseabsw/novelparas.R'), sep = '\n')
```

### numpart.R -- number sections that have generic divider naming

This script is intended to convert dividers `# Part ??` into numbered dividers. It requires generic dividers of the form `* * *` to be converted 
to `# Part ??` with a text editor.

```{r comment='numpart.R', echo=FALSE}
cat(readLines('../baseabsw/numpart.R'), sep = '\n')
```

### splitnovel.R -- divide book text file into individual sections with headers

```{r comment='splitnovel.R', echo=FALSE}
cat(readLines('../baseabsw/splitnovel.R'), sep = '\n')
```

### joinmp3.R -- consolidate, convert and add images

`joinmp3.R` works out which files to consolidate to yield a 
small number of rougly equal-sized mp3 files. These are 
converted to m4a files which are renamed to m4b files. 
Then cover images are added to both sets of files.

```{r comment='joinmp3.R', echo=FALSE}
cat(readLines('../baseabsw/joinmp3.R'), sep = '\n')
```


## Appendix C: List of voices usable by epub2tts-edge

```{r comment='voices', echo=FALSE}
cat(readLines('edgevoices.txt'), sep = '\n')
```
