---
title: "Experiments in generating audiobook files with epub2tts-edge"
author: "John C Nash"
date: "2024-10-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Initial version 2024-07-29, revised 2024-10-19

## Background

Audiobooks have become a popular way to hear rather than read the content of
books, in addition to being an important resource for the visually impaired.
Unfortunately, creating good-quality audiobook files is not easy. Written content
has many features of typography, layout, illustration and idiosynchratic expression
that do not carry over to the spoken word. Moreover, most 
readers, even professional actors,  have difficulty maintaining their enunciation and 
providing expressiveness over the generally extended duration of a book.
Furthermore, the services of professional actors and readers are expensive relative
to the resources of the visually impaired and amateur authors.

The development of large language models for converting text to speech (i.e. TTS)
has reached the stage where it could be considered a possibility for carrying 
out the task of generating audiobooks from textual form, for example, epub 
files. This article is a set of notes on some experiments with freely available
software to do precisely this.

Popularly such software is called "artificial intelligence", but "large language
models" more correctly describes the programs. In particular, the software used
in developing this report comes from the Github project **epub2tts-edge** of
Christopher Aedo. See https://github.com/aedocw/epub2tts-edge. Aedo has done a
great service in this and the earlier **epub2tts** project which is based on 
the Coqui AI TTS tools (https://github.com/coqui-ai/TTS). Moreover, he has 
responded quickly and helpfully to queries.

There are undoubtedly other tools that I could have used, but experimentation is
very time consuming, and preliminary experiments showed promise, first with 
**epub2tts**, then somewhat better results with **epub2tts-edge**. This led to
more, and more ambitious, experiments, as well as some investigations into ways
to improve outputs or make them more usable to potential listeners.

## Source works

I have generally put my creative writing on the UK writers' site *https://obooko.com.*
This currently is twelve novels and two collections of memoir/short story
material. A feature of several of my works is that they involve more than one 
language. My first novel, **Thursday Afternoon**, for example, has a scene where
a young Canadian airforceman is working with some British soldiers and a Flemish
woman to find and clear a landmine in 1944 Flanders. The Flemish woman communicates
with the Canadian in French, but falls back into Flemish on occasions. There are
thus three languages mixed in a couple of pages of text, which is a challenge to
the software. Moreover, I have yet to discover how to inform the software of the
language to be read, though I feel certain there must be a way to do this.

## Installation of the software

Setup followed instructions in the github repositories to install the software for the two 
projects. Note that this may involve a considerable download of code and
binaries depending on the existing software installation. I worked with Linux Mint. 
One limitation for me is a lack of
CUDA capable graphical processing units (GPUs), so performance of the
tools on my computers was somewhat lacklustre. Typically rendering of
an epub to an m4b or mp3 took over an hour.

Note that the `epub2tts-edge` software currently calls for Python support using
at least version 3.11 or later. The audio rendering also needs an active internet
connection. 

The software also was developed for a Linux environment. While there is the possibility
in Windows of running the Windows Subsystem for Linux, I have chosen to use Oracle
VirtualBox to install Linux Mint 22 (Wilma). Indeed, because my working platform is
an earlier Linux Mint and I do not wish to disturb some stable facilities, I have
installed LM22 under my existing version. My writing colleague and coordinator of the
Stittsville Creative Writers Group, R J Partington, installed LM22 under Windows.
This has the advantage of isolating the text to speech work from other ongoing 
activities.

## Standard usage of the software

We assume we have followed instructions in the Github repository to install the
software. 

**epub2tts-edge** is intended to be applied to an **epub** file. (https://www.w3.org/publishing/epub3/, 
https://en.wikipedia.org/wiki/EPUB). Such 
files are packaged collections of HTML text and metadata. Suppose that
we have a file **mybook.epub**. Then we first convert this file to a properly
structured plain text file **mybook.txt** by opening a terminal (command line
screen) and issuing the command

```
       epub2tts-edge mybook.epub
```

This takes only a few seconds to execute. 
This creates a file mybook.txt and extracts a cover image 
mybook.png from the epub file. The text file begins with two 
lines of the form

```
Title: Dodging the Potholes
Author: John C. Nash
```

It also has some lines that commence with a hash symbol (#) that 
give "chapters" or "sections" and the text behind the hash is read
out in the output audio as a chapter heading. However, the algorithm
by which these lines are created is not clear to me, but they can be
artificially inserted into the file, or if generated ones like

```
# Part 24 
```

they can be edited to a more desirable phrase.
Text following the hash is read
as the title of the section. Below we discuss how to break up the work and have separate
audio files, which also serves to allow for correction of individual parts.

### Failure of the epub to text process

We have observed some cases where `epub2tts-edge` fails to carry out the conversion
to text and extraction of the cover image. There are two other approaches we have used.

1) **calibre** program `ebook-convert`

This tool is part of the **calibre** ebook management software. We simply use the
command

```
ebook-convert myfile.epub myfile.txt
```

Generally `myfile.txt` will be more or less correctly formatted, though we have
noticed that the initial "Title:" and "Author:" may be combined in one line. This
can, of course, be corrected with a text editor.

2) **pandoc**

```
pandoc -f epub -t plain -o myfile.txt myfile.epub
```

will create a text version of the epub. Unfortunately, it often breaks paragraphs
into multiple lines which cause undesirable pauses in the eventual audio file(s).
Moreover, the section delimiters ("# (some text)") are often missing. This is the
difficult workaround when all else fails.

### Conversion to audio file(s)

The much longer process is converting the text file to an audio file of type
m4b (https://www.lifewire.com/m4b-file-2621958). For a novel, this may take of the 
order of an hour if the work is
in a single computer text file. Breaking up the text file into parts is therefore
helpful for repairing glitches in individual sections, and my experience is that
these glitches are inevitable. 

The second stage run of **epub2tts-edge** generally requires us to specify a "voice" 
or "speaker". If no speaker is given, a neutral English-only speaker with a quite
moderate American accent is used, which can be specified by the identifier **en-US-AndrewNeural**.
Moreover, for multilingual texts we need to use one of a few special multilingual voices.
For example, the specifier **en-US-AndrewMultilingualNeural** will use the same voice
as the default **for English**, but process multiple languages using **different** speaker
voices for the other languages, which is disconcerting for the listener. Note that it seems 
that different "speaker" specifications will handle the same text differently and some 
words pronounced correctly by one voice will be mangled by another.

The second stage command can also specify a cover image which can be in jpg or png format.
The command is

```
    epub2tts-edge --speaker speakerID --cover mybook.png  mybook.txt
```

where speakerID is a valid speaker identifier and mybook.png (or some other png
or jpg image file) 
is a cover image available in the same directory as mybook.txt. The output will be
a file mybook-speakerID.m4b. This can be a very large file of several hundred megabytes
for a full novel, and take 8-16 hours to play. The m4b format allows bookmarks so the
user can pause and restart. On the other hand, older devices may not be able to play
these files. They are, however, convertible to other audio file formats, in particular
mp3. 

Note that line breaks in the text file mybook.txt cause quite long pauses in the
voice reading. Therefore line breaks that are non-pausing should be removed, giving
full paragraphs on a single "line". 

## Some important general problems

### Omitted text

**WARNING**: I have discovered that text in an epub file that is in a monospaced
(i.e., typewriter-like) font is somehow eliminated completely from the file mybook.txt
and must be re-inserted manually. This is a nuisance and can cause grief. For example,
in my novels, I use such fonts for "letters" and other correspondence or quotations
such as the text on a tombstone. I have had some email correspondence with Christopher
Aedo about this, but so far we have not found the source of the glitch. (?? Note 2024-11-01:
This may have been fixed and needs a test.)

It is feasible to replace the first stage run of epub2tts-edge with a different 
translator from epub to text as we have noted above. Some other choices are noted
here.

- I could not get the software **epub2txt2** (https://github.com/kevinboone/epub2txt2) to 
compile (and hence was unable to install it).

- **epub2txt** is a Python program (https://github.com/ffreemt/epub2txt) that can be
easily installed with the command `pip install epub2txt` if Python is properly set
up. This seems to get all the material, but gives no spaces between what would be
"paragraphs". Moreover, verbatim material has line endings, so some attention to 
that detail may be relevant.

- **pandoc**, as noted above, does not keep paragraphs together as a single "line".
(See 
https://unix.stackexchange.com/questions/647686/convert-epub-to-txt-and-preserve-original-formatting).
`pandoc` leaves two line endings between "paragraphs" but has line endings 
within paragraphs to make the
text easy to display in a text editor that does not wrap lines. It should be 
noted that epub2tts-edge
then pauses for each line ending, so these single line endings need to be removed. 
However, there are
some parts of the epub such as the title and cover pages where we would like 
to retain the single line
endings (or double them to ensure separation of important fields that 
should be read separately).
Note that omitting the "-t plain" option leaves some tags within the 
resulting text file that 
could cause unwanted spoken material. 

- The `ebook-convert` program from the **calibre** software, as noted above, works
reasonably well. It should be the first alternative to `epub2tts-edge` when that
program fails to convert the epub successfully.

### Lost features

I have not been able to get strikeout, bold or italic to be expressed in 
the audio output.

Emphasis, by way of capitalization, is also lost, as can be seen by processing the file 
**cap4emphasis.txt** through epub2tts-edge. Note line endings have been removed.

```
# Capitals for emphasis

Putting words in capitals to emphasize them will only get them pronounced letter by letter in some cases. Hence capitalized "ARE" is not emphasized. Separated letters gives A R E. Simple "are" does not get emphasized. It would be useful to be able to ask **epub2tts-edge** to give emphasis.

"I said," often has awkward pause or mis-emphasis
```

### Difficulties with language change

When the language of the text changes, particularly with dialogue, there can be 
unwelcome results:

- Sometimes the language model, which appears to be linked into the speaker voice chosen,
either fails to recognize the language change, so continues in the current language
setting inappropriately, or else will continue after the "new" language in that language
when it is not correct.

- The voice of the speaker may change drastically in the new language, which is 
disconcerting for the reader.

I have not (at 2024-8-5) yet found work-arounds for these issues.


## Workarounds

Some glitches in pronunciation can be overcome by use of phonetic text. Here
are a number of examples, all of which are small files that can be tested using

```
  epub2tts-edge [file].txt
```

where `[file]` is replaced by the appropriate identifier. Note that in some cases
the long lines needed by epub2tts-edge give display problems for the Rmarkdown
editor in R-studio (https://www.rstudio.com/categories/rstudio-ide/) 
and the resulting pdf output.


### 0pauses.txt

```
0pauses.txt

Various things cause pauses.

Line endings.

Like
these
ones.

A period will give a break. Just like that. But shorter than line ending.

But dashes -- seem to be ignored.

What about a semicolon; does that work? More or less. Perhaps dashes are same.
```

### titles.txt

```
# titles.txt

Lt. Cmdr. Try full words "Lieutenant Commander" or phonetically Leftenant Commander

Putting M I 9 together gets MI9.  Better separated.

Ms. (not sounding like Miz). Try Miz.

Mrs. could be better MISSUS or Missus. Note that epub2tts-edge seems to get this to work OK.

We can also try versus and vs.

Monsieur Lefebvre not M. Lefebvre

POWs  ("s" not pronounced sometimes). Try P O W's. 
```

### titles.txt

```
# names.txt

Vaudreuil try  vohdroy
A / C or Ae slash C for air conditioning or alternating current. Spell it out.
Arnprior Arnpryor Arnprye-ore
Belisle  Belleel
Jean Godin use zjonn gohdann
Lebreton Lebretton
Levis try Levee
Long Sault  (pronunciation)  Long Soo
saltpetre  saltpeter
Paule try Pohl

Almonte use Almont
arboretum
Aylmer
Boucherville probably OK
Dieppe 
Domenico Nacci or Nachi
Erminia
escapees
escheat
Frieda
Greenock -- seems OK
I-81 
Jocelyn
kinesiology and phys. ed. or phonetic fiz ed
Leica 
Maniac
Model A Ford
Myrna pronounced strangely by female voice default of epub2tts
Ontario
Rideau pronounced Rydoh in epub2tts
Rimouski
Tremblay mispronounced  try T R O M B L A Y  Tromblay
V I M Y or Vimy pronounced Veemy or Vy me? By using V I M M Y we get Vimmy.
```


### homonymns.txt

```
# homonyms.txt

"read" sometimes reed, not red

bough of a tree, bow of a ship, a ribbon in a bow.

bowed down

learned vs learnt

row -- pronounced as rou not roe

lead pronounced as "leed" (led for weight)
```

### numberpunctuation.txt

```
#numberpunctuation.txt

Numbers, units and punctuation give difficulties.

2/3 versus two thirds

.2 or point 2 is just 2 with pause. Not Point 2

Brackets (), [], and {} are not handled well

web addresses are often mangled

Lower case M M may be millimeters. M M M is mmm.
35mm vs 35 mm
No. 2  not number 2
cwt. versus hundredweight
```

### other.txt

```
# other.txt

en-US-AndrewNeural speaker handles these correctly
escapees
steno for stenographer
`unmaking` not pronounced correctly in epub2tts
```

### Dialect

Sometimes we want to write dialogue with the speaker using local pronunciation.
Particular cases may be the dropping of the letter H by some working-class
British speakers. This can give problems, and need phonetic replacement.

```
# dialect.txt

I'll 'ave the recording of Ave Maria.

I'll avv the recording of Ave Maria.
```

### Vocalizations
Vocalizations like "oh", "er", "hmmph", "hmmm" sometimes get mangled, though 
epub2tts-edge with the en-US-AndrewNeural voice seems to work reasonably well. 

```
# vocalizations.txt

epub2tts-edge does not seem to know how to properly enunciate some vocalizations common in human speech.

The surprise expressive "oh" is often mis-said. "Oh, I really must fix that."

Disdain, as with "hmmph". Hesitation as with -- er -- and -- hmmm -- or realization with "ah" are potentially problematic.

```

## Audio post-processing

It is sometimes necessary to post-process the audio files.

### Conversion to mp3

Since many older devices cannot play m4b files, but are intended for mp3 files, 
we may need to convert. Unfortunately, most such devices do NOT allow the user
to save the position of a paused play. Therefore it is useful to have an audiobook
in mp3 form in multiple smaller files. One way to do this is to build the audiobook
in pieces, and this is the recommended approach. Nevertheless, **epub2tts-edge**
produces files as m4b files, and these need conversion. I do this with **ffmpeg**
and batch process them with a small (Linux / bash) command-line script. 

```
#!/bin/bash
#
# convm4b
mkdir newfiles
for f in *.m4b; do ffmpeg -i "$f" -codec:v copy -codec:a libmp3lame -q:a 2 newfiles/"${f%.m4b}.mp3"; done
```

**Open question** What is Windows / Mac equivalent?

### Breaking a large mp3 file into pieces

I have also had to break
up a large file (converted from the m4b of an entire novel). For this I used the
**audacity** software which is available on most common platforms. This is a graphical
user interface program allowing for sections of an audio file to be cut and pasted
and exported in various formats, including mp3. 

### Concatenating audio files

Sometimes the mp3 files are really too small, so the user needs to keep loading
them. In that case we want to concatenate several files. **ffmpeg** can do this
using a text file containing a list of the files to be processed in the order they appear.
An example command is 

```
ffmpeg -f concat -i mylist.txt -c copy output.wav
```

which will produce a file of type wav. 

File mylist.txt must have lines of the form 

```
file 'myfile.type' 
```

where type is one of acceptable types to ffmpeg, which does not include
m4b unfortunately, even though we used ffmpeg above to do conversions.

Another tool is **mp3wrap**, which uses a command

```
mp3wrap [options]  OUTPUTFILE  f1.mp3  f2.mp3  [f3.mp3]
```

Unfortunately, there is a limit to the number of files that can be included,
but the number does not appear to be documented. 

I was unable to get the program **mp3cat** from  https://github.com/dmulholl/mp3cat
to work at all. 

### Changing speed or tempo of audio

**audacity** allows the speed or tempo (pitch may change with one of these) to
be altered. For some texts and some "speakers" (i.e., voices) a slowdown of 
approximately 8% made the material easier to comprehend.


## Special Tools

To process the main source text for the second-stage of epub2tts-edge, I have 
written several small scripts to perform particular tasks. I have written these
in **R** or in the Linux **bash** script, but those choices were purely for
convenience. The tasks are relatively straightforward, but tedious to carry out
without the automation.

### Section numbering

Most of my books do not have regular chapters, but do have sections. 
These are typically, but not always, delimited by "* * *".
I use a text editor and replace these (and any other suitable breakpoints)
with  

```
# Part ??
```

An R script numpart.R (see Appendices) is then used (after editing filename specifications
within the script) to auto-replace the "??" with suitable numbers. Note that we
want these numbers to all be the same length for processing to audio files, 
so leading zeros are strongly recommended in the numbering, and the script
prepares such numbers.

The section dividers provided by epub2tts-edge can also be edited
to provide useful output. I have used lines like

```
# The title of this book
```

or 

```
# Capsule comment
```

with the latter for the "teaser" comment for the book.

### Section splitting

It is useful to have individual files of text for each section of a book to
convert to audiobook form. Such files are smaller and more quickly processed, 
so errors or glitches can be repaired and the files re-processed. File 
**splitnovel.R** in the Appendices does this.

Notes:

- we may wish to consolidate several sections after splitting so the audio
playback files are of a reasonable size. This can be done with script **combpart.R**
in the Appendices.  Titles for the consolidated sections could be titled withing 
the text source for conversion by adding title lines in the form

```
# Module A of My Excellent Book
```


- there are issues with sections that are named versus 'Part 043' style, in that
mixing these can be awkward. A suggestion is to use a style

```
# Part 043 -- The plot thickens
```


## A recipe

### Extracting the cover and the text

**epub2tts-edge** aims to do this in the first pass using a command like

`epub2tts-edge mybook.epub`

which should give the cover image in file mybook.png and "text" 
in file mybook.txt. The first lines of mybook.txt, which can be copied to a file 
mybookheader.txt, are something like

```
Title: My fabulous book
Author: Me A. Writer
```

Moreover, the software will try to create "parts" denoted with the hash (#) symbol
followed by a title, but this will depend to a large extent on the epub content
structure.

Unfortunately, as we have noted, **epub2tts-edge**, using the Python library 
**Beautiful Soup** (according to Christopher Aedo), fails to extract
any material in a monospaced font.

As an alternative process, we can use **pandoc** as follows:

- A command like the following will extract text

`pandoc -f epub -t plain -o mybookp.txt mybook.epub`

This file will NOT have headers for parts, nor will it have the header of the 
type mybookheader.txt mentioned above, but the header can be added by using
a text editor. Similarly, whatever divisions are present can be edited to 
the form `# Part ??`, or such divisions can be inserted. Blank lines above
and below the division are recommended.

- The program **epub-cover** can be used to extract the cover image. This is 
part of the Ruby package epub-parser from https://gitlab.com/KitaitiMakoto/epub-parser.

`epub-cover mybook.epub` 

will output whatever cover file is embedded in the epub, which may have a name
somewhat different from the epub title, and may be a JPEG file. Note that one
may choose to use epub2tts-edge to get mybook.png and then ignore the text file
produced in case it is missing material that was in monospaced font.

### Cleanup of pandoc text output

The text file produced by pandoc does not have the mybookheader.txt. Moreover, its
paragraphs are NOT separated suitably for the second (main) stage of epub2tts-edge.

- Add newlines after unseparated lines as appropriate, particularly in title page(s), where we do not want 
paragraph run-on when we process the mybookp.txt to consolidate paragraphs.

### Section dividers

- For my own books, with `* * *` section separator, convert "* * *" to "# Part ??"
throughout. This can be done with most text editors.

- Check that there is a line ending for the final line of the mybook.txt file.

- Add section dividers for the title page(s), the capsule comment and any
preamble. Optionally expand or diminish the section divisions.

### Avoiding some awkward pauses or mispronunciations

In the text file mybook.txt.

- Remove periods after initials in author name(s) to avoid unnecessary pause.

- Remove the copyright symbol (©), if necessary expanding to the word "copyright".

- Put spaces in email addresses e.g. `j.smith@isp.org` becomes `j . s m i t h @ i s p . o r g`.
  Similarly expand abbreviations and postal codes so they are not read as if they are words.

- Expand province or state abbreviations. 

### Consolidate paragraphs

- Run the program **novelparas.R** to ensure text paragraphs are consolidated
for the audio output. Even if the text has been extracted with epub2tts-edge,
there may be manually inserted monospace text that requires paragraph consolidation
to remove unwanted line endings.

### Phonetic corrections

Proper names, dialect, homonyms and similar issues for text to speech should
be corrected. 

- We need a set of source / replacement choices. We shall suppose this is in the
form of lines

```
old (tab)  new
```

This assumes 1-1 choices. We will manually deal with situations where there is 
more than one choice using a text editor. A prototype script to process the
substitutions, offering the user a choice to accept the suggested change (there
are situations where we do NOT wish to take the suggestion), is **wordsub.R**.
This script uses a named file **phoneticsub.txt** which has entries described
above. To save computing time, it is worthwhile having a "master" file, call
is **phoneticsub.full.txt** with all possible replacements. This file can be
copied to phoneticsub.txt and unnecessary entry lines simply removed.

### Section numbering

Run **numpart.R** to get sections numbered.

### Section splitting

Run **splitnovel.R** to divide the source text into individual sections. This is
helpful later when or if we discover that the conversion of text to speech has
made an error or glitch. We can then attempt to correct the issue for just the
single section. Once the text is adjusted appropriately, we can then re-work
the section, if appropriate first inserting it into the consolidated module
where it belongs.

### Section consolidation

To provide suitably sized audio files, we consolidate the sections into modules
using **combpart.R**. 

### Conversion to speech

The script **runedge.R** is intended to process the text modules (or sections)
and produce files of type m4b. Note that this script is currently set up to be
first edited to provide the right filename root for the modules and the correct
cover image. 

This stage of the process takes quite a long time -- typically of the order of
an hour. 

### (Optional) Adjust cover image

The program **AtomicParsley** (https://github.com/wez/atomicparsley) can be
used to insert a different cover image into m4b (and other audio) files.
One syntax is

```
cp myfile.m4b tmp.m4b
AtomicParsley tmp.m4b --artwork REMOVE_ALL
AtomicParsley tmp.m4b --artwork cover.jpg -o tmpx.m4b
mv tmpx.m4b myfile.m4b
rm tmp.m4b
```

### Convert audio files to desired format

While m4b files are the current recommendation for audiobooks, many player
devices do not handle them. Therefore, we may wish to convert the files. The
Linux **bash** script **convm4b.sh** will handle this nicely, but there are
many tools available in the common operating systems. (I welcome suggestions.)

### (Optionally) Adjust audio file characteristics

The program **audacity** is useful to change speed, tempo or pitch of audio
files. Some text-to-speech models create output that is very quick. Slower
and lower-pitched output may be helpful to listeners.

## Open needs

- A fix for the first stage of epub2tts-edge so that monospaced font material
is included. An alternative would be tools to work with other epub to text
translation tools such as **pandoc** or **epub2txt** mentioned above that also 
establish the paragraph structure without line endings. At the moment a text 
editor is used.

- Until and unless the large language models can cope with various proper names,
homonymns and dialect, tools to streamline the replacement of selected words with
phonetic substitutes will be welcome. It would be helpful to have a GUI editor
that highlights target words and offers replacements. Such tools already exist for
spelling checkers, so this is not an idle possibility.

- epub2tts-edge needs an active internet connection to function. There are a number
of situations where it would be important to be able to prepare audio output from
text WITHOUT the need for a connection to the Microsoft service. Moreover, it is
by no means certain that this service will be continued, or continued on the same
basis as at present.

## Special cases 

An attempt to convert the epub of novel "Moonbeam Shadows" (file: Moonbeam.epub)
to text with

epub2tts-edge Moonbeam.epub

failed with the messages as follows.

```
Namespace(sourcefile='Moonbeam.epub', speaker='en-US-AndrewNeural', cover=None)
Cover image saved to Moonbeam.png
Exporting Moonbeam.epub to Moonbeam.txt
Traceback (most recent call last):
  File "/home/john/.local/bin/epub2tts-edge", line 8, in <module>
    sys.exit(main())
  File "/home/john/.local/lib/python3.10/site-packages/epub2tts_edge/epub2tts_edge.py", line 388, in main
    export(book, args.sourcefile)
  File "/home/john/.local/lib/python3.10/site-packages/epub2tts_edge/epub2tts_edge.py", line 113, in export
    author = book.get_metadata("DC", "creator")[0][0]
IndexError: list index out of range
john@M21:~/Videos/moonbeam$ 
```

Similarly, **Dodging Potholes** gave the same error. There are likely others.

## Future goals and needs

1) need a script to run pandoc to text

pandoc -f epub -t plain -o filename.txt filename.epub

2) Need a script to run all the parts
 
   - copy epub to shortened name
   - build profile to use for inputs
   - modify scripts to work together
   - timers and log files
   - testing and reporting.


## Conclusions

This article underlines the reality that we are still in the early stages of
large-scale text to speech, at least in the open-source world. 

- We see a large number of small but important "glitches" in the text-to-speech
engines. While not mentioned much in the article, the Coqui AI TTS was tried,
but found wanting.

- Workarounds have been found for almost all the glitches discovered. These have,
however, requires quite a lot of work. 

- Pronunciation and emphasis issues still are found. 

- The lack of a way to tell the TTS engine what language a segment of text is
using is a serious fault and causes a number of issues. It would also be useful
to be able to specify emphasized or softer speech, or to specify the "speaker"
voice for novels with more than one narrator, such as 
https://www.obooko.com/free-historical-fiction-books/dodging-the-potholes .

- Some books require multiple voices. This is relatively easy to arrange if the
text blocks are broken up, but tags to allow the "speaker" to be changed would
be helpful.

- The **edge** TTS engine require the computer processing text to be internet-
connected. This is a serious obstacle for some applications. It also suggests 
that users may be vulnerable to being held hostage by Microsoft should they
suddenly impose a charging model on the parts of edge-TTS now open.


## Appendix 1: epub2tts-edge voices

?? How to get a list

### How to get audio samples

The program **get-edge-tts-samples.py** shows how to get examples of audio for
different edge-tts speakers.

```
#!/usr/bin/env python3

"""
Basic example of edge_tts usage.
"""

import asyncio
import edge_tts

NAMES = ['en-US-AnaNeural', 'en-US-AndrewNeural', 'en-US-AriaNeural', 'en-US-AvaNeural', \
 'en-US-BrianNeural', 'en-US-ChristopherNeural', 'en-US-EmmaNeural', 'en-US-EricNeural', \
 'en-US-GuyNeural', 'en-US-JennyNeural', 'en-US-MichelleNeural', 'en-US-RogerNeural', \
 'en-US-SteffanNeural', 'de-DE-FlorianMultilingualNeural', 'de-DE-SeraphinaMultilingualNeural',\
 'en-US-AndrewMultilingualNeural', 'en-US-AvaMultilingualNeural', 'en-US-BrianMultilingualNeural',\
 'en-US-EmmaMultilingualNeural', 'fr-FR-RemyMultilingualNeural', 'fr-FR-VivienneMultilingualNeural']

TEXT = "This is a longer string that I am sending to text to speech, and using the python module directly."

async def amain() -> None:
    """Main function"""
    for name in NAMES:
        output = f"sample-{name}.wav"
        print(f"Saving {output}")
        communicate = edge_tts.Communicate(TEXT, name)
        await communicate.save(output)


if __name__ == "__main__":
    loop = asyncio.get_event_loop_policy().get_event_loop()
    try:
        loop.run_until_complete(amain())
    finally:
        loop.close()

```

There are a number of lists of edge-tts voices. One is at https://github.com/rany2/edge-tts/blob/master/README.md


## Appendix A: Possible troubles

The following text MAY still give problems and need to be checked

```
# Oddbods.txt

Nene -- This is an early Rolls Royce jet engine using a centrifugal complressor of the
         Whittle style. It was used in the De Havilland Vampire and a few other early
         jet aircraft.  Better to use Neen to get the right pronunciation.
Sterling currency expression 17/6 is better written 17 and 6. 
ha'penny  as hah penny, not haypenny
"wan" --> W A N 
Marthe -- more like Martha than Mart
Feltham  "ham"
French mis-pronounced. "687 25th Avenue"
Paule, Lise (more like Liss)
souper  
“Sacrément!”
dommage
compris
cintres
“C’est un très grand réfrigérateur.”
tarte au sucre
erable
Houle
Decarie
Ste Catherine
Outrement
Ultramontes
Universit'e
maudit
Anges
Sinterklaas
Bolduc
Pieds nus dans l’aube
Sault Ste Marie,   Regina
Metis
Mon Dieu
V-1 may be better as V 1
```

## Appendix B: Software tools associated with this article

### convm4b.sh -- convert m4b files to mp3

```{r comment='', echo=FALSE}
cat(readLines('convm4b.sh'), sep = '\n')
```

### combpart.R -- combine individual book sections 

The purpose of this script is to create texts that will have a reasonable audio file
duration, particularly when in mp3 form, which is more difficult in most cases to
position or resume playing. The number of parts to combine is set at `np <- 10` in
the present version.

```{r comment='', echo=FALSE}
cat(readLines('combpart.R'), sep = '\n')
```

### novelparas.R -- remove single line endings to get paragraphs

**epub2tts-edge** will pause on line endings. Paragraphs are therefore assumed
to be terminated by a line ending.

```{r comment='', echo=FALSE}
cat(readLines('novelparas.R'), sep = '\n')
```

### numpart.R -- number sections that have generic divider naming

This script is intended to convert dividers `# Part ??` into numbered dividers.
It requires generic dividers of the form `* * *` to be converted to `# Part ??`
with a text editor.

```{r comment='', echo=FALSE}
cat(readLines('numpart.R'), sep = '\n')
```

### splitnovel.R -- divide book text file into individual sections with headers

```{r comment='', echo=FALSE}
cat(readLines('splitnovel.R'), sep = '\n')
```


<!-- ./Thuraft/numparts.R -- number the parts in Coffee Klatch Curmudgeon -->
<!-- ./dodging/numparts.R -- number the parts in Dodging the Potholes -- probably buggy -->
<!-- ./dodging/numpart2.R -- number the parts in the DodgingX.txt file -->
<!-- ./dodging/consolidate/makeffmpeginput.R -->
<!-- ./dodging/splitname.R -- split DodgingY.txt into parts for Anna, Martin and Michelle -->
<!-- 	and put these in separate directories for processing to audio files -->
<!-- ./dodging/splitpart.R -- earlier splitter -- not used -->
<!-- ./curmudgeon/splitpartck.R -- split parts of curmudgeon to have separate files -->
<!-- ./splitter.R -- generic splitter -->
<!-- ./partnums.R -- generic numbering program -->
<!-- ./ontheshelf/partnums.R -- on the shelf numbering -->
<!-- ./ontheshelf/splitterots.R -- split on the shelf (is that needed?) -->
<!-- ./Phoenix/numparts.R -- number Phoenix parts -->

<!-- ### command line scripts -->

<!-- ./edgeinstallepub2tts.sh -- installer for epub2tts-edge -->
<!-- ./dodging/consolidate/convm4b.sh -- convert m4b files to mp3 -->
<!-- ./dodging/consolidate/newfiles/fjoin.sh -- illustrate ffmpeg concatenation.  -->
<!-- ./dodging/martin/martin.sh -- convert Martin voice files to m4b (defines voice) -->
<!-- ./dodging/anna/annasonia.sh -- ditto Anna -->
<!-- ./dodging/anna/done/anna.sh -- ditto Anna (?? Clara voice) -->
<!-- ./dodging/michelle/michelle.sh -- ditto Michelle -->
<!-- ./curmudgeon/convm4b.sh -- as above (to mp3) -->
<!-- ./curmudgeon/makeck.sh -- apply epub2tts-edge to each of the parts of Curmudgion -->
<!-- ./convm4b.sh -- as above m4b to mp3 -->
<!-- ./installepub2tts.sh -- install Aedo's epub2tts (uses Coqui TTS) -->
<!-- ./ontheshelf/tryepub2tts/rx.sh -- OTS trial using Coqui on one part -->
<!-- ./ontheshelf/tryepub2tts/runots.sh -- OTS trial using Coqui on parts -->
<!-- ./ontheshelf/runots.sh -- similar -->
<!-- ./ontheshelf/wrk/runots.sh -- similar -->
<!-- ./ontheshelf/runotsx.sh -- similar (Coqui form is in parts) -->
<!-- ./txt2mp3.sh -- test of gtts-cli on a small piece of text -->
<!-- ./smalltests/e.sh -- small test on e.txt -->
<!-- ./moonbeam/mbinstr.sh  -- make Moonbeam audio -->
<!-- ./Leftoversedge/Leftovers-epub2tts-wav/cwav2mp3.sh Wav to mp3 -->
<!-- ./Leftoversedge/cwav2mp3.sh -- ditto -->
<!-- ./LoveIrontries/ex.sh -- run some parts of Love and Iron -->
<!-- ./CoquiTTS_XTTS_Examples/runAllSpeakers.sh -- Aedo example of all speakers -->
<!-- ./CoquiTTS_XTTS_Examples/xttsSpeaker.sh -- Aedo examples of commands for epub2tts -->


<!-- ## Appendix C: epub2tts voices -->

<!-- **epub2tts** a 1-step tool, but works best with a plain text file, including -->
<!-- the file mybook.txt above. -->

<!-- ``` -->
<!-- epub2tts --speaker (valid Coqui speaker id) --cover mybook.png mybook.txt -->
<!-- ``` -->

<!-- The Coqui speaker id's are different from those in Microsoft's **edge** -->
<!-- tts tools. They are ID's like p340.  -->

<!-- A list of voices can be obtained by ...?? -->

<!-- ## Appendix D: Using a custom voice -->

<!-- I have not had any success in doing this.  -->

<!-- epub2tts --xtts TA01.wav,TA03.wav,TA23.wav --language 'en' LeftoversCalibre.txt -->

<!-- failed after a lot of work with error -->

<!-- File "/home/john/.local/lib/python3.10/site-packages/deepspeed/ops/op_builder/cpu/no_impl.py", line 21, in load -->
<!--     raise ValueError("This op had not been implemented on CPU backend.") -->
<!-- ValueError: This op had not been implemented on CPU backend. -->

## Appendix C: List of voices usable by epub2tts-edge

### Multilingual voices

Name: de-DE-FlorianMultilingualNeural
Gender: Male

Name: de-DE-SeraphinaMultilingualNeural
Gender: Female

Name: en-US-AndrewMultilingualNeural
Gender: Male

Name: en-US-AvaMultilingualNeural
Gender: Female

Name: en-US-BrianMultilingualNeural
Gender: Male

Name: en-US-EmmaMultilingualNeural
Gender: Female

Name: fr-FR-RemyMultilingualNeural
Gender: Male

Name: fr-FR-VivienneMultilingualNeural
Gender: Female

### Single language voices


Name: af-ZA-AdriNeural
Gender: Female

Name: af-ZA-WillemNeural
Gender: Male

Name: am-ET-AmehaNeural
Gender: Male

Name: am-ET-MekdesNeural
Gender: Female

Name: ar-AE-FatimaNeural
Gender: Female

Name: ar-AE-HamdanNeural
Gender: Male

Name: ar-BH-AliNeural
Gender: Male

Name: ar-BH-LailaNeural
Gender: Female

Name: ar-DZ-AminaNeural
Gender: Female

Name: ar-DZ-IsmaelNeural
Gender: Male

Name: ar-EG-SalmaNeural
Gender: Female

Name: ar-EG-ShakirNeural
Gender: Male

Name: ar-IQ-BasselNeural
Gender: Male

Name: ar-IQ-RanaNeural
Gender: Female

Name: ar-JO-SanaNeural
Gender: Female

Name: ar-JO-TaimNeural
Gender: Male

Name: ar-KW-FahedNeural
Gender: Male

Name: ar-KW-NouraNeural
Gender: Female

Name: ar-LB-LaylaNeural
Gender: Female

Name: ar-LB-RamiNeural
Gender: Male

Name: ar-LY-ImanNeural
Gender: Female

Name: ar-LY-OmarNeural
Gender: Male

Name: ar-MA-JamalNeural
Gender: Male

Name: ar-MA-MounaNeural
Gender: Female

Name: ar-OM-AbdullahNeural
Gender: Male

Name: ar-OM-AyshaNeural
Gender: Female

Name: ar-QA-AmalNeural
Gender: Female

Name: ar-QA-MoazNeural
Gender: Male

Name: ar-SA-HamedNeural
Gender: Male

Name: ar-SA-ZariyahNeural
Gender: Female

Name: ar-SY-AmanyNeural
Gender: Female

Name: ar-SY-LaithNeural
Gender: Male

Name: ar-TN-HediNeural
Gender: Male

Name: ar-TN-ReemNeural
Gender: Female

Name: ar-YE-MaryamNeural
Gender: Female

Name: ar-YE-SalehNeural
Gender: Male

Name: az-AZ-BabekNeural
Gender: Male

Name: az-AZ-BanuNeural
Gender: Female

Name: bg-BG-BorislavNeural
Gender: Male

Name: bg-BG-KalinaNeural
Gender: Female

Name: bn-BD-NabanitaNeural
Gender: Female

Name: bn-BD-PradeepNeural
Gender: Male

Name: bn-IN-BashkarNeural
Gender: Male

Name: bn-IN-TanishaaNeural
Gender: Female

Name: bs-BA-GoranNeural
Gender: Male

Name: bs-BA-VesnaNeural
Gender: Female

Name: ca-ES-EnricNeural
Gender: Male

Name: ca-ES-JoanaNeural
Gender: Female

Name: cs-CZ-AntoninNeural
Gender: Male

Name: cs-CZ-VlastaNeural
Gender: Female

Name: cy-GB-AledNeural
Gender: Male

Name: cy-GB-NiaNeural
Gender: Female

Name: da-DK-ChristelNeural
Gender: Female

Name: da-DK-JeppeNeural
Gender: Male

Name: de-AT-IngridNeural
Gender: Female

Name: de-AT-JonasNeural
Gender: Male

Name: de-CH-JanNeural
Gender: Male

Name: de-CH-LeniNeural
Gender: Female

Name: de-DE-AmalaNeural
Gender: Female

Name: de-DE-ConradNeural
Gender: Male

Name: de-DE-FlorianMultilingualNeural
Gender: Male

Name: de-DE-KatjaNeural
Gender: Female

Name: de-DE-KillianNeural
Gender: Male

Name: de-DE-SeraphinaMultilingualNeural
Gender: Female

Name: el-GR-AthinaNeural
Gender: Female

Name: el-GR-NestorasNeural
Gender: Male

Name: en-AU-NatashaNeural
Gender: Female

Name: en-AU-WilliamNeural
Gender: Male

Name: en-CA-ClaraNeural
Gender: Female

Name: en-CA-LiamNeural
Gender: Male

Name: en-GB-LibbyNeural
Gender: Female

Name: en-GB-MaisieNeural
Gender: Female

Name: en-GB-RyanNeural
Gender: Male

Name: en-GB-SoniaNeural
Gender: Female

Name: en-GB-ThomasNeural
Gender: Male

Name: en-HK-SamNeural
Gender: Male

Name: en-HK-YanNeural
Gender: Female

Name: en-IE-ConnorNeural
Gender: Male

Name: en-IE-EmilyNeural
Gender: Female

Name: en-IN-NeerjaExpressiveNeural
Gender: Female

Name: en-IN-NeerjaNeural
Gender: Female

Name: en-IN-PrabhatNeural
Gender: Male

Name: en-KE-AsiliaNeural
Gender: Female

Name: en-KE-ChilembaNeural
Gender: Male

Name: en-NG-AbeoNeural
Gender: Male

Name: en-NG-EzinneNeural
Gender: Female

Name: en-NZ-MitchellNeural
Gender: Male

Name: en-NZ-MollyNeural
Gender: Female

Name: en-PH-JamesNeural
Gender: Male

Name: en-PH-RosaNeural
Gender: Female

Name: en-SG-LunaNeural
Gender: Female

Name: en-SG-WayneNeural
Gender: Male

Name: en-TZ-ElimuNeural
Gender: Male

Name: en-TZ-ImaniNeural
Gender: Female

Name: en-US-AnaNeural
Gender: Female

Name: en-US-AndrewMultilingualNeural
Gender: Male

Name: en-US-AndrewNeural
Gender: Male

Name: en-US-AriaNeural
Gender: Female

Name: en-US-AvaMultilingualNeural
Gender: Female

Name: en-US-AvaNeural
Gender: Female

Name: en-US-BrianMultilingualNeural
Gender: Male

Name: en-US-BrianNeural
Gender: Male

Name: en-US-ChristopherNeural
Gender: Male

Name: en-US-EmmaMultilingualNeural
Gender: Female

Name: en-US-EmmaNeural
Gender: Female

Name: en-US-EricNeural
Gender: Male

Name: en-US-GuyNeural
Gender: Male

Name: en-US-JennyNeural
Gender: Female

Name: en-US-MichelleNeural
Gender: Female

Name: en-US-RogerNeural
Gender: Male

Name: en-US-SteffanNeural
Gender: Male

Name: en-ZA-LeahNeural
Gender: Female

Name: en-ZA-LukeNeural
Gender: Male

Name: es-AR-ElenaNeural
Gender: Female

Name: es-AR-TomasNeural
Gender: Male

Name: es-BO-MarceloNeural
Gender: Male

Name: es-BO-SofiaNeural
Gender: Female

Name: es-CL-CatalinaNeural
Gender: Female

Name: es-CL-LorenzoNeural
Gender: Male

Name: es-CO-GonzaloNeural
Gender: Male

Name: es-CO-SalomeNeural
Gender: Female

Name: es-CR-JuanNeural
Gender: Male

Name: es-CR-MariaNeural
Gender: Female

Name: es-CU-BelkysNeural
Gender: Female

Name: es-CU-ManuelNeural
Gender: Male

Name: es-DO-EmilioNeural
Gender: Male

Name: es-DO-RamonaNeural
Gender: Female

Name: es-EC-AndreaNeural
Gender: Female

Name: es-EC-LuisNeural
Gender: Male

Name: es-ES-AlvaroNeural
Gender: Male

Name: es-ES-ElviraNeural
Gender: Female

Name: es-ES-XimenaNeural
Gender: Female

Name: es-GQ-JavierNeural
Gender: Male

Name: es-GQ-TeresaNeural
Gender: Female

Name: es-GT-AndresNeural
Gender: Male

Name: es-GT-MartaNeural
Gender: Female

Name: es-HN-CarlosNeural
Gender: Male

Name: es-HN-KarlaNeural
Gender: Female

Name: es-MX-DaliaNeural
Gender: Female

Name: es-MX-JorgeNeural
Gender: Male

Name: es-NI-FedericoNeural
Gender: Male

Name: es-NI-YolandaNeural
Gender: Female

Name: es-PA-MargaritaNeural
Gender: Female

Name: es-PA-RobertoNeural
Gender: Male

Name: es-PE-AlexNeural
Gender: Male

Name: es-PE-CamilaNeural
Gender: Female

Name: es-PR-KarinaNeural
Gender: Female

Name: es-PR-VictorNeural
Gender: Male

Name: es-PY-MarioNeural
Gender: Male

Name: es-PY-TaniaNeural
Gender: Female

Name: es-SV-LorenaNeural
Gender: Female

Name: es-SV-RodrigoNeural
Gender: Male

Name: es-US-AlonsoNeural
Gender: Male

Name: es-US-PalomaNeural
Gender: Female

Name: es-UY-MateoNeural
Gender: Male

Name: es-UY-ValentinaNeural
Gender: Female

Name: es-VE-PaolaNeural
Gender: Female

Name: es-VE-SebastianNeural
Gender: Male

Name: et-EE-AnuNeural
Gender: Female

Name: et-EE-KertNeural
Gender: Male

Name: fa-IR-DilaraNeural
Gender: Female

Name: fa-IR-FaridNeural
Gender: Male

Name: fi-FI-HarriNeural
Gender: Male

Name: fi-FI-NooraNeural
Gender: Female

Name: fil-PH-AngeloNeural
Gender: Male

Name: fil-PH-BlessicaNeural
Gender: Female

Name: fr-BE-CharlineNeural
Gender: Female

Name: fr-BE-GerardNeural
Gender: Male

Name: fr-CA-AntoineNeural
Gender: Male

Name: fr-CA-JeanNeural
Gender: Male

Name: fr-CA-SylvieNeural
Gender: Female

Name: fr-CA-ThierryNeural
Gender: Male

Name: fr-CH-ArianeNeural
Gender: Female

Name: fr-CH-FabriceNeural
Gender: Male

Name: fr-FR-DeniseNeural
Gender: Female

Name: fr-FR-EloiseNeural
Gender: Female

Name: fr-FR-HenriNeural
Gender: Male

Name: fr-FR-RemyMultilingualNeural
Gender: Male

Name: fr-FR-VivienneMultilingualNeural
Gender: Female

Name: ga-IE-ColmNeural
Gender: Male

Name: ga-IE-OrlaNeural
Gender: Female

Name: gl-ES-RoiNeural
Gender: Male

Name: gl-ES-SabelaNeural
Gender: Female

Name: gu-IN-DhwaniNeural
Gender: Female

Name: gu-IN-NiranjanNeural
Gender: Male

Name: he-IL-AvriNeural
Gender: Male

Name: he-IL-HilaNeural
Gender: Female

Name: hi-IN-MadhurNeural
Gender: Male

Name: hi-IN-SwaraNeural
Gender: Female

Name: hr-HR-GabrijelaNeural
Gender: Female

Name: hr-HR-SreckoNeural
Gender: Male

Name: hu-HU-NoemiNeural
Gender: Female

Name: hu-HU-TamasNeural
Gender: Male

Name: id-ID-ArdiNeural
Gender: Male

Name: id-ID-GadisNeural
Gender: Female

Name: is-IS-GudrunNeural
Gender: Female

Name: is-IS-GunnarNeural
Gender: Male

Name: it-IT-DiegoNeural
Gender: Male

Name: it-IT-ElsaNeural
Gender: Female

Name: it-IT-GiuseppeNeural
Gender: Male

Name: it-IT-IsabellaNeural
Gender: Female

Name: ja-JP-KeitaNeural
Gender: Male

Name: ja-JP-NanamiNeural
Gender: Female

Name: jv-ID-DimasNeural
Gender: Male

Name: jv-ID-SitiNeural
Gender: Female

Name: ka-GE-EkaNeural
Gender: Female

Name: ka-GE-GiorgiNeural
Gender: Male

Name: kk-KZ-AigulNeural
Gender: Female

Name: kk-KZ-DauletNeural
Gender: Male

Name: km-KH-PisethNeural
Gender: Male

Name: km-KH-SreymomNeural
Gender: Female

Name: kn-IN-GaganNeural
Gender: Male

Name: kn-IN-SapnaNeural
Gender: Female

Name: ko-KR-HyunsuNeural
Gender: Male

Name: ko-KR-InJoonNeural
Gender: Male

Name: ko-KR-SunHiNeural
Gender: Female

Name: lo-LA-ChanthavongNeural
Gender: Male

Name: lo-LA-KeomanyNeural
Gender: Female

Name: lt-LT-LeonasNeural
Gender: Male

Name: lt-LT-OnaNeural
Gender: Female

Name: lv-LV-EveritaNeural
Gender: Female

Name: lv-LV-NilsNeural
Gender: Male

Name: mk-MK-AleksandarNeural
Gender: Male

Name: mk-MK-MarijaNeural
Gender: Female

Name: ml-IN-MidhunNeural
Gender: Male

Name: ml-IN-SobhanaNeural
Gender: Female

Name: mn-MN-BataaNeural
Gender: Male

Name: mn-MN-YesuiNeural
Gender: Female

Name: mr-IN-AarohiNeural
Gender: Female

Name: mr-IN-ManoharNeural
Gender: Male

Name: ms-MY-OsmanNeural
Gender: Male

Name: ms-MY-YasminNeural
Gender: Female

Name: mt-MT-GraceNeural
Gender: Female

Name: mt-MT-JosephNeural
Gender: Male

Name: my-MM-NilarNeural
Gender: Female

Name: my-MM-ThihaNeural
Gender: Male

Name: nb-NO-FinnNeural
Gender: Male

Name: nb-NO-PernilleNeural
Gender: Female

Name: ne-NP-HemkalaNeural
Gender: Female

Name: ne-NP-SagarNeural
Gender: Male

Name: nl-BE-ArnaudNeural
Gender: Male

Name: nl-BE-DenaNeural
Gender: Female

Name: nl-NL-ColetteNeural
Gender: Female

Name: nl-NL-FennaNeural
Gender: Female

Name: nl-NL-MaartenNeural
Gender: Male

Name: pl-PL-MarekNeural
Gender: Male

Name: pl-PL-ZofiaNeural
Gender: Female

Name: ps-AF-GulNawazNeural
Gender: Male

Name: ps-AF-LatifaNeural
Gender: Female

Name: pt-BR-AntonioNeural
Gender: Male

Name: pt-BR-FranciscaNeural
Gender: Female

Name: pt-BR-ThalitaNeural
Gender: Female

Name: pt-PT-DuarteNeural
Gender: Male

Name: pt-PT-RaquelNeural
Gender: Female

Name: ro-RO-AlinaNeural
Gender: Female

Name: ro-RO-EmilNeural
Gender: Male

Name: ru-RU-DmitryNeural
Gender: Male

Name: ru-RU-SvetlanaNeural
Gender: Female

Name: si-LK-SameeraNeural
Gender: Male

Name: si-LK-ThiliniNeural
Gender: Female

Name: sk-SK-LukasNeural
Gender: Male

Name: sk-SK-ViktoriaNeural
Gender: Female

Name: sl-SI-PetraNeural
Gender: Female

Name: sl-SI-RokNeural
Gender: Male

Name: so-SO-MuuseNeural
Gender: Male

Name: so-SO-UbaxNeural
Gender: Female

Name: sq-AL-AnilaNeural
Gender: Female

Name: sq-AL-IlirNeural
Gender: Male

Name: sr-RS-NicholasNeural
Gender: Male

Name: sr-RS-SophieNeural
Gender: Female

Name: su-ID-JajangNeural
Gender: Male

Name: su-ID-TutiNeural
Gender: Female

Name: sv-SE-MattiasNeural
Gender: Male

Name: sv-SE-SofieNeural
Gender: Female

Name: sw-KE-RafikiNeural
Gender: Male

Name: sw-KE-ZuriNeural
Gender: Female

Name: sw-TZ-DaudiNeural
Gender: Male

Name: sw-TZ-RehemaNeural
Gender: Female

Name: ta-IN-PallaviNeural
Gender: Female

Name: ta-IN-ValluvarNeural
Gender: Male

Name: ta-LK-KumarNeural
Gender: Male

Name: ta-LK-SaranyaNeural
Gender: Female

Name: ta-MY-KaniNeural
Gender: Female

Name: ta-MY-SuryaNeural
Gender: Male

Name: ta-SG-AnbuNeural
Gender: Male

Name: ta-SG-VenbaNeural
Gender: Female

Name: te-IN-MohanNeural
Gender: Male

Name: te-IN-ShrutiNeural
Gender: Female

Name: th-TH-NiwatNeural
Gender: Male

Name: th-TH-PremwadeeNeural
Gender: Female

Name: tr-TR-AhmetNeural
Gender: Male

Name: tr-TR-EmelNeural
Gender: Female

Name: uk-UA-OstapNeural
Gender: Male

Name: uk-UA-PolinaNeural
Gender: Female

Name: ur-IN-GulNeural
Gender: Female

Name: ur-IN-SalmanNeural
Gender: Male

Name: ur-PK-AsadNeural
Gender: Male

Name: ur-PK-UzmaNeural
Gender: Female

Name: uz-UZ-MadinaNeural
Gender: Female

Name: uz-UZ-SardorNeural
Gender: Male

Name: vi-VN-HoaiMyNeural
Gender: Female

Name: vi-VN-NamMinhNeural
Gender: Male

Name: zh-CN-XiaoxiaoNeural
Gender: Female

Name: zh-CN-XiaoyiNeural
Gender: Female

Name: zh-CN-YunjianNeural
Gender: Male

Name: zh-CN-YunxiNeural
Gender: Male

Name: zh-CN-YunxiaNeural
Gender: Male

Name: zh-CN-YunyangNeural
Gender: Male

Name: zh-CN-liaoning-XiaobeiNeural
Gender: Female

Name: zh-CN-shaanxi-XiaoniNeural
Gender: Female

Name: zh-HK-HiuGaaiNeural
Gender: Female

Name: zh-HK-HiuMaanNeural
Gender: Female

Name: zh-HK-WanLungNeural
Gender: Male

Name: zh-TW-HsiaoChenNeural
Gender: Female

Name: zh-TW-HsiaoYuNeural
Gender: Female

Name: zh-TW-YunJheNeural
Gender: Male

Name: zu-ZA-ThandoNeural
Gender: Female

Name: zu-ZA-ThembaNeural
Gender: Male


